A hazards approach to the biometric analysis of infant mortality

### 2023-03-30 Slides/slides.Rmd
educational part complete

>>>>

## Demonstration of the B-P plot

Assume 100 births, of which 90 survives infancy, ten death ages:

2, 6, 10, 15, 25, 41, 90, 120, 180, 309.

```{r bpex, fig.height = 5}
par(lwd = 2)
C <- 365 / (log(365 + 1))^3
g <- function(x) C * (log(x + 1))^3
d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
h <- seq_along(d) / 100
plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Deaths / 100", 
     ylim = c(0, 0.12), xlim = c(0, 365))
axis(2, las = 1)
abline(h = 0, v = c(0, 365))
axis(1, at = g(d), labels = d)
axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
box()
abline(v = g(28), lty = 2, col = "magenta")
text(30, 0.11, "Neonatal", col = "red")
text(110, 0.11, "Post-neonatal", col = "blue")
points(g(c(41, 309)), c(6/100, 10/100), col = "red", lwd = 2, pch = 4)
```

Calculation (no ties): $\frac{1}{100}, \frac{2}{100}, \frac{3}{100}, \ldots, \frac{10}{100}.$

<!--
\begin{equation*}
p_n = \frac{n}{100}, \; n = 1, \ldots 10
\end{equation*}
-->

## Demonstration of the B-P plot II

Assume 100 births, of which 90 survives infancy, ten death ages:

2, 6, 10, 15, 25, 41, 90, 120, 180, 309.

```{r bpex2, fig.height = 5}
par(lwd = 2)
C <- 365 / (log(365 + 1))^3
g <- function(x) C * (log(x + 1))^3
d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
h <- seq_along(d) / 100
plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Deaths / 100", 
     ylim = c(0, 0.11), xlim = c(0, 365))

abline(h = 0, v = c(0, 365))
axis(1, at = g(d), labels = d)
axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
box()
abline(v = g(28), lty = 2, col = "magenta")
text(g(12), 0.08, "Exogenous", col = "blue", cex = 0.8)
text(g(12), 0.03, "Endogenous", col = "red", cex = 0.8)
points(g(c(41, 309)), c(6/100, 10/100), col = "blue", lwd = 2)
##
fy <- function(x, xc, yc){
    yc[1] + (yc[2] - yc[1]) / (xc[2] - xc[1]) * (x - xc[1])
}
x1 <- g(41)
y1 <- 6/100
x2 <- g(309)
y2 <- 10/100
x0 <- 0
x3 <- g(309)
y0 <- fy(x0, c(x1, x2), c(y1, y2))
y3 <- fy(x3, c(x1, x2), c(y1, y2))
lines(c(x0, x3), c(y0, y3), lty = 2, lwd = 1)
axis(2, las = 1, at = c(0, y0, y3), labels = round(c(0, y0, y3), 3))
lines(c(0, g(309)), c(y3, y3), lty = 2, lwd = 1)
lines(c(0, 0), c(0, y0), col = "red", lwd = 3)
lines(c(0, 0), c(y0, y3), col = "blue", lwd = 3)
```

Calculation (no ties): $\frac{1}{100}, \frac{2}{100}, \frac{3}{100}, \ldots, \frac{10}{100}.$

<!--
\begin{equation*}
p_n = \frac{n}{100}, \; n = 1, \ldots 10
\end{equation*}
-->

## The hazards plot

Assume 100 births, of which 90 survives infancy, ten death ages:

2, 6, 10, 15, 25, 41, 90, 120, 180, 309 (same as before).

```{r hazex, fig.height = 5}
par(lwd = 2)
C <- 365 / (log(365 + 1))^3
g <- function(x) C * (log(x + 1))^3
d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
h <- seq_along(d) / (100:91)
plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Cumulative hazards", 
     ylim = c(0, 0.12), xlim = c(0, 365))
h2 <-  seq_along(d) / 100
lines(g(d), h2, type = "b", lty = 3, lwd = 1) 
axis(2, las = 1)
abline(h = 0, v = c(0, 365))
axis(1, at = g(d), labels = d)
axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
box()
abline(v = g(28), lty = 2, col = "magenta")
text(g(12), 0.11, "Neonatal", col = "red")
text(g(110), 0.11, "Post-neonatal", col = "blue")
ttt <- sum(g(c(41, 90, 120, 180, 309))) + 90 * (g(365) - g(28))

```

Calculation (no ties): $\frac{1}{100}, \frac{2}{99}, \frac{3}{98}, \ldots, \frac{10}{91}.$

## The hazards plot II

<!-- Assume 100 births, of which 90 survives infancy, ten death ages:

2, 6, 10, 15, 25, 41, 90, 120, 180, 319 (same as before).
-->

```{r hazex2, fig.height = 5}
par(mai = c(1,1,1,1), lwd = 2)
C <- 365 / (log(365 + 1))^3
g <- function(x) C * (log(x + 1))^3
d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
h <- seq_along(d) / (100:91)
plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Cumulative hazards", 
     ylim = c(0, 0.13), xlim = c(0, 365))

axis(2, las = 1)
abline(h = 0, v = c(0, 365))
axis(1, at = g(c(d, 365)), labels = c(d, 365))
tabb <- g(c(0, 28, 41, 90,120, 180, 309, 365))
axis(3, at = tabb, labels = round(tabb))
box()
abline(v = g(28), lty = 2, col = "magenta")
##text(g(12), 0.08, "Neonatal", col = "red")
text(g(130), 0.11, "Total")
text(g(130), 0.05, "Exogenous", col = "blue")
ttt <- sum(g(c(41, 90, 120, 180, 319) - 28)) + 90 * (g(365) - g(28))
lambda <- 5 / ttt
x <- g(0:365)
y <- Hweibull(x, shape = 1, scale = 1 / lambda)
lines(x, y, lty = 2, col = "blue")
axis(4, at = c(0, y[366], h[10]), labels = round(c(0, y[366], h[10]), 3), las = 1)
lines(c(365, 365), c(0, y[366]), col = "blue", lwd = 3)
lines(c(365, 365), c(y[366], h[10]), col = "red", lwd = 3)
text(30, 0.12, "Neonatal", col = "red", cex = 0.8)
text(110, 0.12, "Post-neonatal", col = "blue", cex = 0.8)
```

Exogenous slope = $\frac{\#\text{(post deaths)}}{\text{post exposure}} = \frac{5}{27626} \approx 0.00018$

Post exposure = $90 (365 - 68) + (93 - 68) + (163 - 68) + \cdots + (335 - 68) \approx 27626$

## Adding the endogenous mortality

```{r hazex3, fig.height = 5}
par(mai = c(1,1,1,1), lwd = 2)
C <- 365 / (log(365 + 1))^3
g <- function(x) C * (log(x + 1))^3
d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
h <- seq_along(d) / (100:91)
plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Cumulative hazards", 
     ylim = c(0, 0.13), xlim = c(0, 365))

axis(2, las = 1)
abline(h = 0, v = c(0, 365))
axis(1, at = g(c(d, 365)), labels = c(d, 365))
tabb <- g(c(0, 28, 41, 90,120, 180, 309, 365))
##axis(3, at = tabb, labels = round(tabb))
box()
abline(v = g(28), lty = 2, col = "magenta")
##text(g(12), 0.08, "Neonatal", col = "red")
text(g(130), 0.11, "Total")
text(g(130), 0.02, "Exogenous", col = "blue")
text(g(130), 0.06, "Endogenous", col = "red")
ttt <- sum(g(c(41, 90, 120, 180, 319) - 28)) + 90 * (g(365) - g(28))
lambda <- 5 / ttt
x <- g(0:365)
y <- Hweibull(x, shape = 1, scale = 1 / lambda)
lines(x, y, lty = 2, col = "blue")
axis(4, at = c(0, y[366], h[10]), labels = round(c(0, y[366], h[10]), 3), las = 1)
lines(c(365, 365), c(0, y[366]), col = "blue", lwd = 3)
lines(c(365, 365), c(y[366], h[10]), col = "red", lwd = 3)
text(30, 0.12, "Neonatal", col = "red", cex = 0.8)
text(110, 0.12, "Post-neonatal", col = "blue", cex = 0.8)
## Add neonatal:
xn <- g(d)
yn <- h - Hweibull(xn, shape = 1, scale = 1 / lambda)
lines(xn, yn, type = "l", col = "red", lty = 4)
```
